---
title: "Analyze EMMAX Results"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(gwasrapidd)
library(tidyverse)
library(ggplot2)
library(data.table)

```

```{r load_data}
source("scripts/utils/manhattan_plot.R")
# All files ending in .ps (EMMAX association output)
files <- list.files(path="output/emmax/", pattern="\\.ps$")
```

Q-Q Plots and Manhattan Plots
-----------------------------

Q-Q plots to check if there is any effect of population substructure on the results, and genomic inflation factor (lambda).

Manhattan plots were generated for all SNPs that had a significant association (p < 0.05).

filtering SNPs that were not genome-wide suggestive (p >= 1e-05) to identify genes that were close to the suggestive SNPs.

```{r}
for (file in files) {
  
  cat("Processing", file, "\n")
  
  gwas.results <- fread(paste0("output/emmax/", file), sep="\t", header=F)
  colnames(gwas.results) <- c("SNP", "Beta", "Beta.SE", "P")
  
  # Calculate genomic inflation factor
  chisq <- qchisq(1 - gwas.results[["P"]], 1)
  lambda <- median(chisq) / qchisq(0.5, 1)
  cat("\t", "Genomic Inflation Factor:", lambda, "\n")

  # QQ Plot
  plot.data <- data.frame(
    Observed.Log.10.P=-log10(sort(gwas.results[["P"]])),
    Expected.Log.10.P=-log10(1:nrow(gwas.results) / (nrow(gwas.results) + 1))
  )
  p <- ggplot(plot.data) +
    geom_point(aes(x=Expected.Log.10.P, y=Observed.Log.10.P), size=I(0.75), alpha=0.5) +
    geom_line(aes(x=Expected.Log.10.P, y=Expected.Log.10.P)) +
    labs(x="Expected -log10(P)", y="Observed -log10(P)") +
    annotate("text", x=6, y=1, label=paste0("λ=", round(lambda, 5))) +
    basic_theme_big

  ggsave(paste0("output/emmax/", strsplit(file, "\\.")[[1]][1], "_qqplot.png"), p, width=8, height=8)
  
  # Filter out all non-significant SNPs (p >= 0.05)
  gwas.results <- as.data.frame(gwas.results[gwas.results[["P"]] < 0.05,])
  
  # Mutate to match manhattan plot helper function requirements
  chr.bp <- do.call(rbind, strsplit(gwas.results$SNP, ":"))
  gwas.results$CHR = as.numeric(chr.bp[,1])
  gwas.results$BP = as.numeric(chr.bp[,2])

  # Manhattan Plot
  q <- manhattan.plot(gwas.results)
  ggsave(paste0("output/emmax/", strsplit(file, "\\.")[[1]][1], "_manhattan.png"), q, width=15, height=6)

  # Filter to suggestive SNPs (p < 1e-05)
  gwas.results <- gwas.results %>%
    dplyr::filter(P < 1e-05) %>%
    dplyr::select(CHRBP=SNP, CHR, BP, Beta, Beta.SE, P) # changed header for "SNP" to match FUMA
  
  write.table(gwas.results, paste0("output/emmax/", strsplit(file, "\\.")[[1]][1], "_suggestive_snps.tsv"), row.names=F, quote=F, sep="\t")
}
```

Annotations
-----------

Variants were annotated using FUMA's SNP2GENE and GENE2FUNC functions (https://fuma.ctglab.nl/).
Resulting files, including parameter configurations, are in output/2b_emmax_summary/FUMA_archetype_*
Minimum P-value of lead SNPs was changed to 1^-5
sample size 562
Otherwise all parameters were set to default.
Main observation: 13-22 top genes. Not very related to brain function. Maybe to connective tissues and circulation 


GWAS Catalog
-------------
Quarying NHGRI-EBI Gwas Cataloue for published associations of mapped variants.
```{r, gwascatalog, warning=FALSE, error=FALSE}
phenotypes <- c("archetype_1", "archetype_2", "archetype_3", "archetype")
pubs <- c()
for (ph in phenotypes) {
  fuma <- read_tsv(paste0("./output/3b_emmax_summary/FUMA_", ph, "/snps.txt"))
  variants <- fuma$rsID
  for (var in variants) {
    st <- get_studies(variant_id=var)
    if (length(st@studies$study_id)>0) {
      pubs <- rbind(pubs,
                    cbind(phenotype=ph, SNP=var, nearest_gene=fuma$nearestGene[fuma$rsID==var], 
                          reported_trait=st@studies$reported_trait, st@publications))
    }
  }
}
write_csv(pubs, path=paste0("./output/3b_emmax_summary/emmax_gwascat.csv"), append = TRUE)

```




