---
title: "3_architecture"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

rm(list=ls())

library(dplyr)
library(ggplot2)
library(data.table)
```

```{r}
pheno <- c("archetype", "archetype_1", "archetype_2", "archetype_3")
# Read in genes from FUMA
loci <- c()
for (ph in pheno) {
    mat <- read.table(paste0("output/2b_emmax_summary/FUMA_", ph, "/genes.txt"), header=T, stringsAsFactors=F)
    loci <- rbind(loci, cbind(mat$symbol, ph, "Subtypes"))
}

# Read in Nikhil's genes associated with clinical phenotypes and submodules, filtering out modules
loci.NM <- read.csv("output/3_architecture/loci_NM.csv", header=T, stringsAsFactors = F)
colnames(loci) <- colnames(loci.NM) <- c("Locus","Phenotype","Type")
loci <- rbind(loci, loci.NM, stringsAsFactors=F) %>%
    dplyr::filter(Type!="Modules")

write.csv(loci, "output/3_architecture/loci.csv", row.names=F, quote=F)
# loci <- read.csv("output/2b_emmax_summary/loci.csv", header=T, stringsAsFactors = F)


# Export for Cytoscape - edge table
# (same as loci table in this case, just different column names)
edge.table <- as.data.frame(table(loci))
edge.table <- edge.table %>% 
  dplyr::filter(Freq != 0) %>%
  dplyr::select(Source=Phenotype, Target=Locus, Type)
write.csv(edge.table, "output/3_architecture/edge.table.csv", row.names=F, quote=F)

# Export for Cytoscape - node table
node.table <- data.frame(
  node_name=c(unique(loci$Locus), unique(loci$Phenotype)),
  node_type=c(rep("Locus", length(unique(loci$Locus))), rep("Phenotype", length(unique(loci$Phenotype))))
)
rownames(node.table) <- node.table$node_name

# Add display name for phenotypes
node.table$display_name <- ifelse(node.table$node_type=="Phenotype", as.character(node.table$node_name), NA)
# Add display name for shared loci (except "archetype")
tab <- dplyr::filter(edge.table, Source!="archetype")
shared <- as.data.frame(table(tab$Target)) %>%
  dplyr::filter(Freq > 1) %>%
  dplyr::select(Var1)
shared <- as.character(shared[[1]])
node.table[shared,"display_name"] <- as.character(node.table[shared,"node_name"])
write.csv(node.table, "output/3_architecture/node.table.csv", row.names=F, na="", quote=F)

```
Notable features of the network:
The archetypes don't overlap at all between them (subtypes don't overlap either between A and B)
subtype A can be broken down into loci associated with archetype 1 vs 2; consistent with exp
subtype B overlap a lot with archetype 3; consistent with exp
archetypes have 3 loci overlap with clinical dx vs 1 for subtypes (2 for arc3, 1 for arch2)
arch3 connects braak and cerad with one loci

Annotating the maped genes:
```{r}
# loci that are shared between phenotypes
#shared.loci <- edge.table %>%
#   filter(duplicated(edge.table$Target) | duplicated( edge.table$Target, fromLast = TRUE)) %>%
#   arrange(Target)

## GWAS catalog results from FUMA
a1 <- read.table("output/2b_emmax_summary/FUMA_archetype_1/gwascatalog.txt", header=TRUE, sep="\t")

```
