---
title: "2b Top genes annotations"
output:
  html_document:
    toc: true
    code_folding: hide
---
This script annotates the top genes associated with each archetype in terms of (sub)modules and GO enrichment, and cell specifity
```{r setup}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, cache=TRUE, cache.extra = R.version)
```

```{r libaries}
suppressPackageStartupMessages({
  library(clusterProfiler)
  library(GOsummaries)
  library(vcd)
  library(tidyverse)
  })
```

# Input
```{r}
topgenes <- readRDS("analyses/rnaseq/1_fit_archetypes/topgenes_logFC_k3p20_AD.RDS")

submodules <- read.csv(file="data/milind2019/submodules.csv", na.strings = "") %>%
  dplyr::filter(Cohort=="rosmap") %>%
  droplevels()
bigmodules <- read.csv(file="data/milind2019/bigmodules.csv", na.strings = "") %>%
  dplyr::filter(Cohort=="rosmap") %>%
  droplevels()

# cell-specific markers from McKenzie et al. 2017
hum.specific <- read.csv("data/McKenzie_2017_Brain_Cell_Specific_Markers_Human_Specificity.csv", header=T, stringsAsFactors=F)

```

# Modules and submodules enrichment
Add information about modules and submodules to top genes  
Add their HGNC.ID as well, consistent with those in NM's submodules
```{r ModuleAnnotation, warnnig=FALSE, message=FALSE}
topgenes <- topgenes %>%
  left_join(select(bigmodules, Ensembl.ID, Module), by = "Ensembl.ID") %>%
  left_join(select(submodules, Ensembl.ID, Module, Submodule), by =c("Ensembl.ID", "Module")) %>%
  select(Ensembl.ID, HGNC.ID, Module, Submodule, everything())

```

How many of the top genes are in one of the modules/submodules?
``` {r compare_modules}
cat(round(100*length(which(!is.na(topgenes$Submodule)))/nrow(topgenes)), "% of topgenes are in one of the submodules", "\n") # 62% in the submodules
cat(round(100*length(which(!is.na(topgenes$Module)))/nrow(topgenes)), "% of topgenes are in one of the modules", "\n") # 84% in the big modules

data <- select(topgenes, Module, Archetype) %>%
  mutate(Module=substring(Module, 6))
mytable <- xtabs(~ Module + Archetype, data)
ftable(mytable) # print table
summary(mytable) # chi-square test of indepedence
assoc(mytable, shade=TRUE, labeling_args = list(abbreviate_labs = c(Module=6), varnames=c(FALSE,FALSE), rot_labels=c(0,0)))
pdf(file = "analyses/rnaseq/2b_topgenes_annotations/chi2_modules.pdf")
assoc(mytable, shade=TRUE,labeling_args = list(abbreviate_labs = c(Module=6), varnames=c(FALSE,FALSE), rot_labels=c(0,0)))
dev.off()

data <- select(topgenes, Submodule, Archetype) %>%
  mutate(Submodule=substring(Submodule, 6))
mytable <- xtabs(~ Submodule + Archetype, data)
ftable(mytable) # print table
summary(mytable) # chi-square test of indepedence
assoc(mytable, shade=TRUE, labeling_args = list(abbreviate_labs = c(Submodule=6), varnames=c(FALSE,FALSE), rot_labels=c(0,0)))
pdf(file = "analyses/rnaseq/2b_topgenes_annotations/chi2_submodules.pdf")
assoc(mytable, shade=TRUE, labeling_args = list(abbreviate_labs = c(Submodule=6), varnames=c(FALSE,FALSE), rot_labels=c(0,0)))
dev.off()

```
arch3_B
yellow_1: trans-synaptic signaling

arch2, A+B
blue_1: astrocyte metabolism; catabolic; oxydation-reduction
blue_4: astrocyte and microglia activation; cellular motility & regeneration
brown_1: gliogenesis; oligodendrocytic myelination and lysosomal activity

arch1, A
blue_2: cytokine (immune) signaling; inflamatory
blue_3: microglial activation; immune response regulation
turquoise_2: protein folding; cellular response to stress

# GO annotations for topgenes

```{r GOannotations}
archetypes <- unique(topgenes$Archetype)

all.go <- c()
for (arch in archetypes) {
    cat("Annotating: ", arch, "\n")
# translate ensembl ID to entrez ID
 eg <- bitr(topgenes[topgenes$Archetype == arch, "Ensembl.ID"], fromType="ENSEMBL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
  
  # Annotate module
  goBP.anno <- enrichGO(gene=eg[,2], ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 0.01, qvalueCutoff=0.05, OrgDb="org.Hs.eg.db", readable=TRUE)
  goBP.anno <- as.data.frame(goBP.anno@result)
  
  # Add to data frame
  if (nrow(goBP.anno) > 0) {
    all.go <- rbind(all.go, cbind("Archetype"=arch, goBP.anno))
  }
}
colnames(all.go)[2] <- "GO.ID"
colnames(all.go)[9] <- "HGNC.ID"

# write out lists to generate tag clouds with REVIGO
for (arch in archetypes) {
  all.go %>%
    dplyr::filter(Archetype==arch) %>%
    dplyr::select(GO.ID, p.adjust) %>%
    write_tsv(paste0("analyses/rnaseq/2b_topgenes_annotations/REVIGO_",arch, ".tsv"))
}

## GOsummaries and cloud plots
gs_arch <- list()
for (arch in archetypes) {
  gs_arch[[arch]] <- topgenes %>%
    dplyr::filter(Archetype==arch) %>%
    dplyr::select(HGNC.ID) %>%
    GOsummaries::gosummaries() %>%
    GOsummaries::add_to_slot.gosummaries("Title", list(arch))
}

for (arch in archetypes) {
  plot(gs_arch[[arch]],filename=paste0("analyses/rnaseq/2b_topgenes_annotations/GOcloud_",arch,".pdf"))
}
```

# Cell specificity for top genes

Following Nikil's 6a_submodule_annotation.Rmd
Cell type markers in the brain are taken from McKenzie et al. (Supplemental File 1, DOI:10.1038/s41598-018-27293-5). The study developed three different metrics to rank the relation of genes to specific cell types (enrichment, expression, specificity). 
The cell specificity for a given gene is the minimum log fold change when pairwise comparing the expression of the gene in the given cell type with the expression of the gene in all other cell types.
Here I'm using cell specificity score to annotate the top genes of each archetype, with chi-square to test for biased representations of any cell type in any archetype, and a plot of deviations from expectation for visualization.

```{r cell_specificity}

archetypes <- sort(unique(topgenes$Archetype))
#NUM.MARKERS = 100

# split topgenes into a list of gene sets by archetype
gene.sets <- split(topgenes, topgenes$Archetype, drop=TRUE) %>%
  lapply(function(l){dplyr::select(l, HGNC.ID)})

cell.types <- c("Astroctyes", "Endothelial", "Microglia", "Neurons", "Oligodendrocytes")

hum.specific <- split(hum.specific, hum.specific$Celltype, drop=TRUE) %>%
  lapply(function(l){dplyr::select(l, HGNC.ID=gene, Grand.Mean=grand_mean)})
names(hum.specific) <- cell.types

# Number of marker genes for each cell type in each archetype gene set
specific.set <- sapply(archetypes, function(s) {
  sapply(cell.types, function(ct) length(intersect(hum.specific[[ct]]$HGNC.ID, gene.sets[[s]]$HGNC.ID)), USE.NAMES=T)
})

names(dimnames(specific.set)) <- c("Cell Type", "Archetype")
chisq.test(specific.set)
assoc(t(specific.set), shade=TRUE,labeling_args = list(abbreviate_labs = c(6, 9)))

pdf(file = "analyses/rnaseq/2b_topgenes_annotations/cellTypes.pdf")
assoc(t(specific.set), shade=TRUE,labeling_args = list(abbreviate_labs = c(6, 9)))
dev.off()
#
```

# Output
```{r, cache=FALSE}
saveRDS(topgenes, "analyses/rnaseq/2b_topgenes_annotations/topgenes.anno.RDS")
saveRDS(all.go, file="analyses/rnaseq/2b_topgenes_annotations/all.GO_archetypes.RDS")
```

# Session info
```{r session-info, cache = FALSE}
devtools::session_info()
```
